Supplementary Materials And Methods

### 2019-09-17 Supplementary_mat_met.Rmd
sup met edit and spell check

# 1. Defined functions for ceRNA models and workflow of method
We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous section. The functions and steps of approach are explained as following (Figure S1) :

**Convertion of dataset**: priming_graph function processes the given dataset that includes competing elements in first variable and repressive element in second variable. If the affinity and/or degradation factors are specified in the function, factors are taken into account, are processed with defaults in vice versa. The formulations that are given in equations (1-4) are performed in this function. This step gives the graph object which contains efficiency values of miRNA:competing target pairs in steady-state in terms of amount. It is assumed that the initial target amounts in the dataset is observed after the reppressive activity of miRNAs in steady-state.

![Workflow for simulation of competing endogenous RNA regulations. Graph object in steps 2-6 is saved and updated continuously.](workflow1.png){#fig1}

@@

**Trigger change in graph**: The dataset are assumed as steady-state in previous step and the efficieny coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for disturbtion of steady-state. To provide the disturbtion in the network the workflow offer two methods: update_variables and update_how. The first, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.

**Updating current values of variables**: After variables updating in edge varibles, these are carried to node variables. Current and previous values of variables are stored as node variables with update_variables function.

@@

The node elements in the dataset are handled as two type; repressives (miRNAs) and competings (targets). It is assumed in approach that while targets are degrading or inhibiting by miRNAs continuously, miRNAs reversibly used. If the trigger of the network is a miRNA, it maintains the current value of amount that provides by user. On the contrary, it tries to help this process to provide steady-state through the regulations on its amount, if a competing element is used as a trigger. The functions that are used in the approach are developed with R programming so as can be used with other packages. These are can be found in the github repository [ceRNAnetsim github page](https://github.com/selcenari/ceRNAnetsim) and improved with contributions of others.

@@

See Figure S1 in Supplementary Tables file.

@@

- 2. We have obtained graph after change on Gene2 expression as followings:

@@

- 3. We have determined regulations after Gene2 Upregulation:

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Sequential iteration of minsampdata", fig.subcap= c("First response of system to Gene2 upregulation (2nd iteration)", "Spreading of perturbation on system (3th iteration)")}

@@

We have made the same analysis in present of interaction factors.

@@

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Sequential iteration of minsampdata with interaction factors", fig.subcap= c("First response of system to Gene2 upregulation (2nd iteration)", "Spreading of perturbation on system (3th iteration)")}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation with interaction factors",2)
```

When the graphs which were resulted from analyses were examined, it was observed that behaviours were same. But, when the results were analysed in terms of expression values, the regulation differences can be observed.

@@

There are hundreds of defined miRNAs for human, so this results in presence of common targets of miRNAs in cells. Therefore, we have analysed perturbation efficiency of common target in *minsamp* dataset.

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Perturbation of Gene4 on minsampdata with interaction factors", fig.subcap= c("1st iteration after Gene4 upregulation", "2nd iteration after Gene4 perturbation")}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene4", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Common Gene-Gene4 Upregulation without interaction factors",2)
```

@@

## Determination of perturbation efficiencies efficiencies of elements in system.

@@

We have obtained the gene expression values of patient using [TCGAbiolinks](https://www.bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html) package from [Bioconductor](https://www.bioconductor.org/). For this process, we have followed the instructions of the package. `TCGAbiolinks` package provides to obtain data for whole number of given barcode(s) at once. But, we preferred to download them seperately to show datasets.

@@

+ The datasets were directly downloaded from supplementary data files of the studies.
+ It was provided that the datasets are converted to same human genome build.
+ The seed type information was organized as the same style.
+ The datasets were combined.
+ We commited the interaction factors as numeric values according to previous studies. (We added the interaction factors and their numeric values at Supplementary tables S2-3)

@@

# ABCC1 gene has 5827 read count in tumor tissue although 1420 in normal tissue (FC=4.10)
```
Secondly, we have determined the most important nodes of network. We applied `find_node_perturbation` function. We only defined nodes that affect the other nodes more than 1.05 fold change with 10 iteration when they increase 3 fold.

@@

The dynamics of the approach are shown in package viggnettes [link](https://selcenari.github.io/ceRNAnetsim/articles/convenient_iteration.html).

@@

Actually, we have developed to provide a new approach mirna mediated regulation networks. This approach may not explain the whole regulation behaviors between miRNAs and targets but can be first step to more detailed and cohorent miRNA:target regulation approach.

>>>>

# 1. Functions defined for ceRNA models and workflow of method
We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous section. The functions and steps of approach are explained as following (Figure S1) :

**Conversion of dataset**: `priming_graph()` function processes the given dataset that includes competing elements in first variable and repressive element in second variable. If the affinity and/or degradation factors are specified in the function, factors are taken into account, are processed with defaults in vice versa. The formulations that are given in equations (1-4) are performed in this function. This step gives the graph object which contains efficiency values of miRNA:competing target pairs in steady-state in terms of amount. It is assumed that the initial target amounts in the dataset is observed after the repressive activity of miRNAs in steady-state.

![Workflow for simulation of competing endogenous RNA regulations. Graph object in steps 2-6 is saved and updated continuously.](workflow1.png){#figs1}

@@

**Trigger change in graph**: The dataset are assumed as steady-state in previous step and the efficiency coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for distribution of steady-state. To provide the distribution in the network the workflow offer two methods: update_variables and update_how. The first, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.

**Updating current values of variables**: After variables updating in edge variables, these are carried to node variables. Current and previous values of variables are stored as node variables with update_variables function.

@@

The node elements in the dataset are handled as two type; repressive (miRNAs) and competing (targets). It is assumed in approach that while targets are degrading or inhibiting by miRNAs continuously, miRNAs reversibly used. If the trigger of the network is a miRNA, it maintains the current value of amount that provides by user. On the contrary, it tries to help this process to provide steady-state through the regulations on its amount, if a competing element is used as a trigger. The functions that are used in the approach are developed with R programming so as can be used with other packages. These are can be found in the Github repository [ceRNAnetsim Github page](https://github.com/selcenari/ceRNAnetsim) and improved with contributions from others.

@@

See Table S1 in Supplementary Tables file.

@@

- 2. We have obtained graph after change on Gene2 expression as following:

@@

- 3. We have determined regulations after Gene2 upregulation:

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2,  fig.show='hold', fig.cap = "Sequential iteration of minsamp data. a) First response of system to Gene2 upregulation (2nd iteration). b) Spreading of perturbation on system (3th iteration)", fig.subcap= NULL}

@@

We have made the same analysis in presence of interaction factors.

@@

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Sequential iteration of minsamp data with interaction factors", fig.subcap= c("First response of system to Gene2 upregulation (2nd iteration)", "Spreading of perturbation on system (3th iteration)")}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation with interaction factors",2)
```

When the graphs which were resulted from analyses were examined, it was observed that behaviors were same. But, when the results were analysed in terms of expression values, the regulation differences can be observed.

@@

Genes targeted by multiple miRNAs (referred to as "common target") are of special interest since they are subject to cooperative effect. Also, they perturb more than one neighborhood. In our small dataset, minsamp, Gene4 is regulated by two miRNAs. Let's simulate perturbation effects triggered by Gene4.

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Perturbation of Gene4 on minsamp data with interaction factors", fig.subcap= c("1st iteration after Gene4 upregulation", "2nd iteration after Gene4 perturbation")}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene4", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Common Gene-Gene4 Upregulation without interaction factors",2)
```

@@

## Determination of perturbation efficiencies of elements in system.

@@

We have obtained the gene expression values of patient using [TCGAbiolinks](https://www.bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html) package from [Bioconductor](https://www.bioconductor.org/). For this process, we have followed the instructions of the package. `TCGAbiolinks` package provides to obtain data for whole number of given barcode(s) at once. But, we preferred to download them separately to show datasets.

@@

+ The datasets were directly downloaded from supplementary data files of the studies.
+ It was provided that the datasets are converted to same human genome build.
+ The seed type information was organized as the same style.
+ The datasets were combined.
+ We committed the interaction factors as numeric values according to previous studies. (We added the interaction factors and their numeric values at Supplementary tables S2-3)

[#TODO çok karmaşık komutlar için ayrı bir klasör hazırlanabilir. burayada link verilebilir: "experimentalmirnagene" tablosu repodaki X.R script ile elde edilmiştir" gibi]

@@

# ABCC1 gene has 5827 read count in tumor tissue although 1420 in normal tissue (FC=4.10)
```

[#TODO SERPINE2]

Secondly, we have determined the most important nodes of network. We applied `find_node_perturbation` function. We only defined nodes that affect the other nodes more than 1.05 fold change with 10 iteration when they increase 3 fold.

@@

The dynamics of the approach are shown in package  [vignettes](https://selcenari.github.io/ceRNAnetsim/articles/convenient_iteration.html).

@@

Actually, we have developed to provide a new approach mirna mediated regulation networks. This approach may not explain the whole regulation behaviors between miRNAs and targets but can be first step to more detailed and coherent miRNA:target regulation approach.
